Phylogenetics and the human microbiome.

### 2014-06-24 reviews/response.txt

> 4) Pg. 15: “However, whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate” – This requires some modification.  Maybe add “one” after “every”.

Done, thanks.

> 5) Pg. 16: “PD under random sub-sampling, which is often applied to enable comparison be- tween samples of different sequencing depths and called “rarefaction curves.”” – Would it be preferable to write that the random subsamples are described (or depicted) with rarefaction curves?

You caught a strange sentence there. I've just cut the strange bits and linked to a later discussion.

@@

> 9 (Kuczynski et al. 2010). However

Thanks. I've added commas.

> P12 “Several groups are continually revising taxonomy...” and P13 “An
> obvious application of phylogenetics is to perform taxonomic
> classification...” We have here two long and technical paragraphs
> (e.g. F-measure, precision and recall, but what is classified and
> how?). You should check this again for clarity and to ensure
> self-contentedness.

You are right. I've decided to just eliminate the technical sentences.

### 2014-06-24 sbreview.tex
Addressing some comments.

\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage{palatino}
\usepackage{wrapfig}
\usepackage[hidelinks]{hyperref}
\usepackage[round]{natbib}
\usepackage[labelfont=bf,labelsep=period,font=small]{caption} % bold the 'Figure #' in the caption and separate it with a period
\newcommand{\forarxiv}[1]{}
\newcommand{\notforarxiv}[1]{#1}

% for SB submission
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
\linespread{1.6}

@@

The parameter regime and focus of human-associated microbial research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
The answer is simple: this system is data- and question-rich.
It absolutely requires molecular methods because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.

@@

Before proceeding, we note that phylogenetics for microbes differs in some important respects from, say, mammalian phylogenetics.
Where phylogenetic research on mammals is converging on one or two possible basic structures for their evolution, much of the deep history of microbes remains obscure.
Inference of this history is complicated by the fact that any tree-like signal in the overall evolution of bacteria is restricted to a small set of so-called ``core'' genes \citep{bapteste2009prokaryotic,Leigh2011-zu,Lang2013-zt}.

@@

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the PhyloCode project \citep[to develop a taxonomic scheme expressed directly in terms of a phylogeny; see][]{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy \citep{kreig1984bergey}, although the actual revision process appears opaque.
The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree, which is annotated as follows.
The first step is to label the internal nodes of the tree to optimize the $F$-measure (the harmonic mean of precision and recall) across clades, where the taxonomic label of a clade is thought of as a prediction for the descendant sequence taxonomic labels.
The second step reassigns sequence taxonomic labels based on these clade taxonomic labels.
Interestingly, tax2tree allows for polyphyletic taxonomic groups as an allowance for either phylogenetic error, lack of resolution of the 16S gene, or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI or EMBL sequence databases.

@@

Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
Unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are commonly used in human microbiome studies, abundance weighted phylogenetic diversity measures are not.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a sublinear function.
\citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiota; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''

@@

\section{Acknowledgements}
I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 MCEB Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and thank the NIH (R01 HG005966-01) and NSF (1223057) for financial support.

>>>>

\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage{palatino}
\usepackage{wrapfig}
\usepackage[hidelinks]{hyperref}
\usepackage[round]{natbib}
\usepackage[labelfont=bf,labelsep=period,font=small]{caption} % bold the 'Figure #' in the caption and separate it with a period
\newcommand{\forarxiv}[1]{#1}
\newcommand{\notforarxiv}[1]{}

% for SB submission
% \renewcommand{\section}[1]{%
% \bigskip
% \begin{center}
% \begin{Large}
% \normalfont\scshape #1
% \medskip
% \end{Large}
% \end{center}}
% \renewcommand{\subsection}[1]{%
% \bigskip
% \begin{center}
% \begin{large}
% \normalfont\itshape #1
% \end{large}
% \end{center}}
% \renewcommand{\subsubsection}[1]{%
% \vspace{2ex}
% \noindent
% \textit{#1.}---}
% \renewcommand{\tableofcontents}{}
% \bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
% \linespread{1.6}

@@

The parameter regime and focus of human-associated microbial research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
The answer is simple: this system is data- and question-rich.
It absolutely requires molecular methods because microbes are now primarily identified by their molecular sequences, and identification of microbes by their molecular sequences is much more straightforward to do in high throughput than morphological or phenotypic characterization.
Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.

@@

Before proceeding, we note that phylogenetics for microbes differs in some important respects from, say, mammalian phylogenetics.
Where phylogenetic research on mammals is converging on one or two possible basic structures for their evolution, much of the deep history of microbes remains obscure.
Inference of this history is complicated by the fact that any tree-like signal in the deep evolutionary relationships between bacteria is restricted to a small set of so-called ``core'' genes \citep{bapteste2009prokaryotic,Leigh2011-zu,Lang2013-zt}.

@@

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the PhyloCode project \citep[to develop a taxonomic scheme expressed directly in terms of a phylogeny; see][]{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy \citep{kreig1984bergey}, although the actual revision process appears opaque.
The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree in a way that allows polyphyletic taxonomic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI or EMBL sequence databases.

@@

Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
The (unweighted and unrooted) PD of a set of taxa $S$ is simply the total of the lengths of the branches which connect taxa in $S$ \citet{faith1992conservation}.
It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
Unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are commonly used in human microbiome studies, abundance weighted phylogenetic diversity measures are not.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a sublinear function.
\citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiota; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling (see discussion of rarefaction below).

@@

\section{Acknowledgements}
I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 MCEB Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript.
This work was supported by National Institutes of health grant R01-HG005966-01 and National Science Foundation grant 1223057.
